`flippant` -- An `R` package for the automated  analysis of dithionite

### 2015-09-19 inst/manuscript/manuscript.rmd
Keep on refining;

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
lipids between the leaflets of lipid bilayers that constitute biological membranes
[@waechter_flippases_2013]. Independent of energy equivalents such as ATP they 
facilitate this ``flipping'' by providing to polar lipid head groups a mechanism
of transition through the hydrophobic environment of membranes, thus reducing 
the massive energetic cost that process would carry otherwise 
[@kornberg_inside-outside_1971]. The underlying molecular mechanisms remain 
largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified **(I was looking for a recent 
citation that clearly spelled that out - in vain. Do you have a proposal?)**. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, we speculate that lipid scrambling may, in fact, 
be an intrinsic property of G-protein coupled receptors in general 
[@goren_constitutive_2014].

To interrogate factors of this functional class, we have developed a biochemical 
assay suited to characterize scramblase activity of purified proteins and protein 
extracts [@chang_chemical_2004] based on the reconstitution of scramblases or 
candidate proteins into preformed synthetic liposomes that contain trace amounts
of nitrobenzoxadiazole (NBD) labeled fluorescent lipids.
Upon addition of membrane-impermeable reducing agents such as dithionite to 
these liposomes, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence disappears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the 
assay has also been used to characterize the stoichiometry of scramblase 
homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the 
package ...**

Starting with the identification of baseline and post-reduction fluorescence 
from fluorometer-generated spectra, via the calculation of the underlying 
statistics and through publication-grade representation of the results, data 
analysis in the context of this assay has until now involved an extensive series
of manual steps, at most supported by spreadsheet calculation facilities
[see the supplementary material of @menon_opsin_2011 for extensive step-wise discussion].
Based on the `R` free software environment for statistical computing and 
graphics [@r_core_team_r:_2015] and prominently drawing on the power and 
elegance of the `ggplot2` plotting system [@wickham_ggplot2_2009], we here 
provide `flippant`, a package that comprehensively, fast and reproducibly covers
all data analysis and graphing needs from raw fluorescence spectra produced by 
scramblase assays. A simple parametrizing spreadsheet is the sole required user 
input.

@@

Based on @menon_opsin_2011 and @goren_constitutive_2014, data is processed as 
follows. Input is format checked and defaults are injected for facultative 
parameters/columns as appropriate. Fluorescence spectra are parsed. This includes 
algorithmic determination of when dithionite was added to the sample using peak
detection by continuous wavelet transformation as implemented in the 
`wmtsa` `R` package [@percival_wavelet_2000;@constantine_wmtsa:_2013]. The 
acquisition time is reset such that the zero time point henceforth corresponds 
to the time of addition. 
Pre-dithionite-addition baseline fluorescence $F_{\text{baseline}}$ is 
determined for each spectrum by averaging (median) over the ten values preceding
dithionite addition. Post-dithinonite-addition minimum fluorescence 
$F_{\text{min}}$ is analogously calculated from the last ten datapoints after 
the default 400 s or the timepoint of measurement supplied by the user in the 
parameter spreadsheet. $F_{\text{min}}$ is volume-corrected based on supplied 
reaction volumes with and without dithionite and for each spectrum/datapoint a 
fluorescence reduction is calculated as follows:
$$
y_{\text{raw}} = 1 - \frac{F_{\text{min}}}{F_{\text{baseline}}}
$$
A relative fluorescence reduction $y$ is calculated scaled to a (required) liposomes-only/no-protein control and the probability $P_{\geq 1}$ for a 
liposome to contain at least one scramblase molecules is calculated using
$$
P_{\geq 1} = \frac{(y - y_0)}{(y_{\text{max}} - y_0)}
$$
where $y$ is the relative fluorescence reduction for the proteoliposome 
measurement and $y_0$ for a measurement of protein-free liposomes. 
$y_{\text{max}}$ refers to either the maximal $y$ in the experiment or is 
derived from a monoexponential fit to the data (default) as a precaution for the 
case where the protein/phospholipid titration did not reach the plateau of the 
saturation curve. The behavior is adjustable using the `scaleTo` parameter of 
the corresponding functions (see below). Next a monoexponential curve is fitted
to the realtionship of $P_{\geq 1}$ to the protein to phospholipid ratio (PPR; 
(mg/mmol)), providing the scramblase-characterizing fit factor $\alpha$.

#Main Features
##Template Generation for Data Input
The following function generates a spreadsheet program-compatible tab-delimited 
`ASCII` template including column names, commentary, expected data type and the 
default used when ommited, supporting the user in providing the data needed.
```{r, code=formatR::usage('scramblaseAssayInputTemplate'), eval=FALSE}
```
The application agnostic tab-delimited format must be maintained for the 
completed table, which is subjected to extensive input checks when read by 
`flippant`.

>>>>

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
lipids between the leaflets of lipid bilayers that constitute biological membranes
[@waechter_flippases_2013]. Independent of energy equivalents such as ATP they 
facilitate this ``flipping'' by providing a mechanism of transition through the 
hydrophobic environment of membranes to polar lipid head groups, thus reducing 
the massive energetic cost that process would carry otherwise 
[@kornberg_inside-outside_1971]. The underlying molecular mechanisms remain 
largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified **(I was looking for a recent 
citation that clearly spelled that out - in vain. Do you have a proposal?)**. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, we speculate that lipid scrambling may, in fact, 
be an intrinsic property of G-protein coupled receptors in general 
[@goren_constitutive_2014].

To interrogate this functional class of proteins, we have developed a biochemical 
assay suited to characterize scramblase activity of purified proteins and protein 
extracts based on the reconstitution of scramblases or candidate proteins into 
preformed synthetic liposomes that contain trace amounts of nitrobenzoxadiazole 
(NBD) labeled fluorescent lipids [@chang_chemical_2004].
Upon addition of membrane-impermeable reducing agents such as dithionite to 
these liposomes, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence disappears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the 
assay has also been used to characterize the stoichiometry of scramblase 
homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].
**Are you aware of further use? More adoption would fortify the need for the 
package ...**

Starting with the identification of baseline and post-reduction fluorescence 
from fluorometer-generated spectra, via the calculation of the underlying 
statistics and through publication-grade representation of the results, data 
analysis in the context of this assay has until now involved an extensive series
of manual steps, at most supported by spreadsheet calculation facilities
[see @menon_opsin_2011;@goren_constitutive_2014 for extensive step-wise discussion].
Based on the `R` free software environment for statistical computing and 
graphics [@r_core_team_r:_2015] and prominently drawing on the power and 
elegance of the `ggplot2` plotting system [@wickham_ggplot2_2009], we here 
provide `flippant`, a package that comprehensively, fast and reproducibly covers
all data analysis and graphing needs from raw fluorescence spectra produced by 
scramblase assays. A simple parametrizing spreadsheet is the sole required user 
input.

@@

Based on @menon_opsin_2011 and @goren_constitutive_2014, data is processed as 
follows. Input is format checked and defaults are injected for facultative 
parameters/columns as appropriate. Fluorescence spectra are parsed. This includes 
algorithmic determination of when dithionite was added to the sample using peak
detection by continuous wavelet transformation as implemented in the 
`wmtsa` `R` package [@percival_wavelet_2000;@constantine_wmtsa:_2013]. 
Acquisition time is aligned for all spectra in the data set such that the zero 
time point henceforth corresponds to the time of addition. 
Pre-dithionite-addition baseline fluorescence $F_{\text{baseline}}$ is 
determined for each spectrum by averaging (median) over the ten values preceding
dithionite addition. Post-dithinonite-addition minimum fluorescence 
$F_{\text{min}}$ is analogously calculated from the last ten datapoints after 
a default 400 s or the timepoint of measurement supplied by the user in the 
parameter spreadsheet. $F_{\text{min}}$ is volume-corrected based on supplied 
reaction volumes with and without dithionite and for each spectrum/datapoint a 
fluorescence reduction is calculated as follows:
$$
y_{\text{raw}} = 1 - \frac{F_{\text{min}}}{F_{\text{baseline}}}
$$
A relative fluorescence reduction $y$ is calculated scaled to a (required) liposomes-only/no-protein control and the probability $P_{\geq 1}$ for a 
liposome to contain at least one scramblase molecules is calculated using
$$
P_{\geq 1} = \frac{(y - y_0)}{(y_{\text{max}} - y_0)}
$$
where $y$ is the relative fluorescence reduction for the proteoliposome 
measurement and $y_0$ for a measurement of protein-free liposomes. 
$y_{\text{max}}$ refers to either the maximal $y$ in the experiment or is 
derived from a monoexponential fit to the data (default) as a precaution for the 
case where the protein/phospholipid titration did not reach the plateau of the 
saturation curve. The behavior is adjustable using the `scaleTo` parameter of 
the corresponding functions (see below). Next a monoexponential curve is fitted
to the realtionship of $P_{\geq 1}$ to the protein to phospholipid ratio (PPR; 
(mg/mmol)), providing the scramblase-characterizing fit factor $\alpha$.

#Main Features
##Template Generation for Data Input
Supporting the user in providing the input data needed, the following function 
generates a spreadsheet program-compatible tab-delimited `ASCII` template, 
including column names, commentary, expected data type and the default used when
ommited.
```{r, code=formatR::usage('scramblaseAssayInputTemplate'), eval=FALSE}
```
The application agnostic tab-delimited format must be maintained for the 
completed table, which is subjected to extensive input checks when read by 
`flippant`.
